Early Warning Signals and the Prosecutor's Fallacy

### 2012-04-23 ReplyToReviewers.md

Dear Tim Holt,

Thank you for handling the reviews of our manuscript.  We are pleased to provide the updated version addressing the issues highlighted by the reviewers and hope that it is ready for publication.  Please find our comments below addressing the points made by each referee.  A pdf document highlighting all changes is also attached.  (Our replies below in _italics_)

Sincerely,

Carl Boettiger

Alan Hastings

@@

Referee: 1
==========

### Comments to the Author
I think that this is one of the most important papers yet written in what is emerging as an area of great interest.  My only concern is that it is written in a highly technical way that may not be accessible to the broad audience that should read it.   The idea of using a model-based approach and approaching the problem as one of model choice is very clever, but I had to read the section three times before I knew what that meant.   I would move some of that material up front to make the main points a little less drily than they are currently made. I think also some technical terms should be embellished with discussion and explanation up front.

_We thank the reviewer for this assessment, and we have done our best to address this important issue, clarifying definitions, expanding the explanation and making other small changes that should make this more accessible._

@@

1. Line 41: “powerful way” is gratuitous

_"powerful" has been deleted_

2. Lines 61-63: I think some transcritical bifurcations can be “blue-sky” transitions; the claim should be qualified

_We apologize that the jargon is may not be used consistently in the field.  Transcritical bifurcations swap the stability of two fixed points, the fixed points to not appear/disappear into the "blue".  We have removed the term_

@@

5. Line 248: “This introduces a nuisance parameter known as the statistical significance criterion” is overly prejudicial. Differences of statistical philosophy are fine, but I think this goes over the line.

_This has been rephrased (we meant only in the technical sense as a parameter you must choose arbitrarily rather than estimate from data, but the point is well taken)_

6. Line 301: Typo

_The missing noun (a citation to Dakos et al. 2008) has been inserted,  thanks for catching this._

### 2012-04-23 boettiger.tex
minor edits

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
that do not depend on estimating the parameters of a known model of a specific system. 
This has led to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Here we extend earlier work by providing estimates 
of the ability of different potential indicators to accurately signal impending regime shifts,
and develop new approaches that both are more efficient
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.  
We distinguish this question from the extensive literature involving change-point analysis 
for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodionov2004, Lenton2009}, 
which is of little use if the goal is the advanced detection of the shift.

We begin by discussing the limitations of current approaches that rely on summary statistics
and provide a description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals. 
We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.
Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
as a way to illustrate the sensitivity of different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.
This approach provides a visualization of the types of errors that arise and how one can trade off between them.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
A saddle-node bifurcation is the result of a stable node and an unstable
saddle-point colliding and annihilating, leaving the system to transition 
to another attractor~\citep[see ][or other dynamical systems text 
for an introduction to this and other bifurcations]{Guckenheimer1983}.

Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches must confront numerous challenges.
In this paper we argue for a model-based approach,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations involve a change in stability of a fixed point, rather
than the sudden dissapearance of a fixed point that has made critical transitions so worrisome.% --
%they are characterized by a steady approach towards the alternative stable state, 
%such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle-node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
must further assume that a changing parameter has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
\citet{Livina2012} indicate that such noise induced transitions cannot be predicted through early warning signals -- 
at least they are not expected to exhibit the same early warning patterns of increased variance and increased autocorrelation
anticipated in the case of a saddle-node bifurcation.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle-node bifurcation, 
the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
A basic difficulty arises from the need to assume a time-series is ergodic: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signal detection,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2012} introduces the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter.
We have added a subscript $t$ to the bifurcation parameter as a reminder that
it is the value which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
Allowing the stochastic term to scale with the square root of $\phi$ follows from the assumption that
of an internal-noise process, such as demographic stochasticity, that arises in deriving the SDE from
a Markov process, see~\citet{Kampen2007a} or~\citet{Black2012}.  
%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order:
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the system is stable, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

\subsection*{Likelihood calculations}
The probability \(P(X|M) \) of the data \(X\) given the model \(M\) is the product of the probability of observing each point in the time series given the previous point and the length of the interval,
\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}

@@

By estimating the stable and critical transition models from the data,
and simulate 500 replicate data sets under each as in the analysis above,
we can then calculate the warning signals statistic over a sliding window
of size equal to one-half the length of the time series,
and compute the correlation coefficient $\tau$ measuring the degree to which the statistic shows an increasing trend.
This results in a distribution of $\tau$ values coming from a model of a stable system,
and a corresponding distribution of $\tau$ values coming from the model with an impending transition.
These distributions are shown in Figure~\ref{fig:summary_box}.
Contrary to the expectation that replicates of the null model (stable system, Equation~\eqref{OU}) would cluster around zero, 
while the test model, Equation~\eqref{LSN}, would cluster around larger positive $\tau$ values, 
the observed $\tau$ values on the replicates extend evenly across the range.  
This results in dramatic overlap and offers little ability to distinguish between the stable replicates 
and the replicates approaching a transition.

@@

\section{Acknowledgments}
We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins and N. Ross for comments on the manuscript. 
This research was supported by funding from NSF Grant EF 0742674 
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
Data and code are available at \href{https://github.com/cboettig/earlywarning}{ https://github.com/cboettig/earlywarning}.

>>>>

Indeed, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
that do not depend on estimating the parameters of a known model of a specific system. 
This has led to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Here we extend earlier work by providing estimates 
of the ability of different potential indicators to accurately signal impending regime shifts,
and develop new approaches that both are more efficient
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.  
We distinguish this question from the extensive literature involving change-point analysis 
for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodionov2004, Lenton2009}, 
which is of little use if the goal is the advanced detection of the shift.

We begin by discussing the limitations of current approaches that rely on summary statistics
and provide a description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals. 
We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.
Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
as a way to illustrate the sensitivity different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.
This approach provides a visualization of the types of errors that arise and how one can trade off between them,
and is important for framing the problem as one focused on prediction.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
A saddle-node bifurcation occurs when a parameter changes and a stable equilibrium 
(node) and an unstable equilibrium (saddle) coalesce and dissapear.  The system
then moves to a more distant equilbrium.  \citet{Guckenheimer1983} or any other 
textbook on dynamical systems will provide precise definitions and further explanation.

Typical patterns used as warning signals include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches must confront numerous challenges.
In this paper we argue for a model-based approach to warning signals,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations involve a change in stability of a fixed point, rather
than the sudden dissapearance of a fixed point that has made critical transitions so worrisome.% --
%they are characterized by a steady approach towards the alternative stable state, 
%such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detects transitions driven by
saddle-node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
must further assume that a changing parameter has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition in system behavior.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes a harvested population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
\citet{Livina2012} indicate that such noise induced transitions cannot be predicted through early warning signals -- 
at least they are not expected to exhibit the same early warning patterns of increased variance and increased autocorrelation
anticipated in the case of a saddle-node bifurcation.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle-node bifurcation, 
the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
A basic difficulty arises from the need to assume a time-series is \emph{ergodic}: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
typical methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signal detection,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2012} introduce the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

\subsection*{Models}
In the neighborhood of a bifurcation a system can be transformed into its \emph{normal form}
by a change of variables to facilitate analysis ~\citep{Guckenheimer1983}.
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter.
We have added a subscript $t$ to the bifurcation parameter as a reminder that
it is the value which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
the system near the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
Allowing the stochastic term to scale with the square root of $\phi$ follows from the assumption that
of an internal-noise process, such as demographic stochasticity, that arises in deriving the SDE from
a Markov process, see~\citet{Kampen2007a} or~\citet{Black2012}. 
The square root could be removed for an external noise process, such as environmental noise.
In practice it will be difficult to descriminate between the square root and linear scaling
in these applications, since the average value of the state changes little before the bifurcation.

%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order:

\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}

Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the underlying system is not changing, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,

\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}

This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

\subsection*{Likelihood calculations}
The probability \(P(X|M) \) of the data \(X\) given the model \(M\) is the product of the probability of observing each point in the time series given the previous point and the length of the interval,

\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}

@@

By estimating the stable and critical transition models from the data,
and simulating 500 replicate data sets under each as in the analysis above,
we can then calculate the warning signals statistic over a sliding window
of size equal to one-half the length of the time series,
and compute the correlation coefficient $\tau$ measuring the degree to which the statistic shows an increasing trend.
This results in a distribution of $\tau$ values coming from a model of a stable system,
and a corresponding distribution of $\tau$ values coming from the model with an impending transition.
These distributions are shown in Figure~\ref{fig:summary_box}.
Contrary to the expectation that replicates of the null model (stable system, Equation~\eqref{OU}) would cluster around zero, 
while the test model, Equation~\eqref{LSN}, would cluster around larger positive $\tau$ values, 
the observed $\tau$ values on the replicates extend evenly across the range.  
This results in dramatic overlap and offers little ability to distinguish between the stable replicates 
and the replicates approaching a transition.

@@

\section{Acknowledgments}
We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins, N. 
Ross, M. Holden, and two anonymous reviewers for comments on the manuscript. 
This research was supported by funding from NSF Grant EF 0742674 
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
Data and code are available at \href{https://github.com/cboettig/earlywarning}{ https://github.com/cboettig/earlywarning}.

### 2012-04-23 fallacy.tex
minor edits

Statisical patterns associated with sudden, catastrophic transitions or ``tipping points''
could play an important role as early warning signs or leading indicators of
these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.
Historical examples of sudden transitions taken from the paleoclimate record
provide an important way to test and evaluate potential leading indicator methods,
and have been widely used for this purpose~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010} 
Similarly, it has been suggested that data gathered on ecological systems such as lakes
that were monitored before the collapsed, or grasslands subjected to overgrazing,
could contain data that could help reveal when similar systems are approaching 
a tipping point~\citet{Inman2011}.  This is a silver lining to the clouds of past
catastrophes -- we may be able to learn from our mistakes.

@@

% Explain variance, autocorrelation, tau.

\emph{Add more context \dots }

@@

 \section{Discussion}
At first it seems tempting to argue that this bias is in fact valuable -- each
of these systems did indeed collapse, so the fact that they show increased 
probability to exhibit warning signals could be taken as a successful detection.
Unfortunately, this reasoning is flawed. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from the sample
that did not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  Taking a closer look at the patterns involved we can
begin to understand why common indicators such as autocorrelation and variance
can be misleading.

Systems that crash by chance excursion beyond the tipping point

Systems collapsing by chance do so from a string of unexpected events

Discuss the pattern of a system crashing by chance -- why this looks like increased autocorrelation and increased variance.

Discuss why this cannt be considered a successful detection.

These are not more likely to crash than another system.

Not a good test of the method's ability to detect bifurcations.

>>>>

Statisical patterns associated with sudden, catastrophic transitions or ``tipping points''
could play an important role as early warning signs or leading indicators of
these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.

% Explain Critical Slowing down, variance, autocorrelation, tau.

% Saddle-node bifurcation
\emph{Add more context \dots }

Historical examples of sudden transitions taken from the paleoclimate record
provide an important way to test and evaluate potential leading indicator methods,
and have been widely used for this purpose~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010} 
Similarly, it has been suggested that data gathered on ecological systems such as lakes
that were monitored before the collapsed, or grasslands subjected to overgrazing,
could contain data that could help reveal when similar systems are approaching 
a tipping point~\citet{Inman2011}.  This is a silver lining to the clouds of past
catastrophes -- we may be able to learn from our mistakes.

@@

@@

 \section{Discussion}

%We have shown ..

It seems tempting to argue that this bias is in fact valuable -- each
of these systems did indeed collapse, so the fact that they show increased 
probability to exhibit warning signals could be taken as a successful detection.
Unfortunately, this reasoning is flawed. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from the sample
that did not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  Taking a closer look at the patterns involved we can
begin to understand why common indicators such as autocorrelation and variance
can be misleading.

% Delete this paragraph?
The system simulated in our examples is governed by a birth-death process --
a discrete Markov process that can in any instant only change by a single 
birth or single death.  At each moment in time, there is some probability 
that either event will occur next.  The system near its stable equilibrium
can be well approximated as a Gaussian process, where consecutive random 
draws buffet it above and below the stable point. While these random draws
vary in magnitude, recall that this is simply an approximation to the 
underlying process where events are all of the same magnitude but vary
in their frequency following a state-dependent Poisson process.  This 
means that what looks like a large random draw from the Gaussian distribution
really corresponds to a series of more rapid events with some excess of 
birth or deaths.

The farther the sytem gets from that stable point (while still 
remaining in the local region), the more likely it is to draw a random step 
that returns it towards the stable point, like a spring that is stretched 
further and further, but there is some probability that it will move further
still. Systems that manage to extend this random walk far enough to cross the
tipping point must do so rather quickly -- with a string of events moving 
closer and closer to the edge.  This pattern, clearly visible before the 
crashes in each of the examples in Figure 1, produces a string of observations
that appear more highly autocorrelated (if we are sampling the system
frequently enought to catch the excursion at all) than we observe in the rest
of the fluctuations around the equilbrium.  Yet this autocorrelation comes from
a chance trajectory moving quickly \emph{away} from the stable state, not from
the critical slowing down pattern in the return times to the stable state
which precede a saddle-node bifurcation and motivate the early warning signal.

This longer than expected excursion results in a higher than expected variance
in that window as well. Both variance and autorcorrelation are calculated using
a moving window over the time-series, which allows the method to pick out a 
pattern of change as the window moves along the sequence. If this chance excursion 
that precedes the crash happens to fill a significant part of the moving window,
the resulting pattern will tend to show an increase in autocorrelation or variance.
If the chance excursion is relatively rapid compared to the frequency at which
the system is observed (spacing of the data) or the width of the moving window, 
the excursion may not significantly alter the general pattern.  In this way, 
some of the events in which a crash is observed will appear to present these
statistical patterns of increased variance or autocorrelation without being
harbingers of approaching critical transitions.

Because we have selected our data by conditioning
on the fact that they have experienced such a crash, these excursions appear
more frequently than we would expect by chance in the observations of a given
stable system over some finite time window.  That biased selection is the
heart of the Prosecutor's Fallacy.  In these cases, the presence of increasing
autocorrelation or increasing variance does not represent the critical slowing
down process.
